Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-09-13 Manuscript.Rmd
Änderungen von Vali übernommen.

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
  - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in N deposition. We considered different drivers of plant community change (i.e. N deposition, climate warming, land-use change) in Swiss mountain hay meadows using data from the Swiss biodiversity monitoring. For this purpose, we compared indicator values of species that disappeared or colonized a site (turnover) with the indicator values of randomly chosen species from the same site. While oligotrophic plant species were more likely to colonize compared to random expectation, we found only weak shifts in plant community structure. Particularly, the average nutrient value of plant communities remained stable over time (2003–2017). Despite this, we found that the nutrient values of colonization species showed largest deviations from random expectation suggesting that N deposition or other factors that change the nutrient content of soils were important drivers of the species composition change over the last 15 years in Swiss mountain hay meadows. In addition, we observed an overall replacement of species with lower indicator values for temperature by species with higher values. Apparently, the community effects of the replacement of eutrophic species with oligotrophic species was outweighed by climate warming. Our results add to the increasing evidence that inferring species turnover will generate a far more reliable understanding of the biotic response to changing environments than solely tracking average community composition.

@@

# Introduction {-}
Nitrogen (N) deposition is the entry of reactive nitrogen compounds into soil, water and vegetation input from the atmosphere to the biosphere. Since nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low nitrogen availability [@Vitousek1997], the addition of nitrogen is likely to alter these communities. Indeed, N deposition together with land-use and climate change is one of the major threats to biodiversity [@Sala2000; @Bobbink2010; @Murphy2013]. While there is strong evidence for the reduction in diversity of species-rich grasslands due to increased N deposition [@Stevens2006; @Dupre2010; @Maskell2010; @Wesche2012], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully diminished emissions of nitrogen oxides, with affiliated decrease in N deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that suffered from increased N deposition. First, N deposition is still high at many sites, since, in contrast to nitrogen oxides, the recent decrease of ammonia emissions was only minor. Thus, even if N deposition is reduced, large areas might still be above the critical load [@Slootweeg2015; @RihmAchermann2016]. Second, there is a potential that communities reach an alternative stable state after decades of increased N deposition and these species are unlikely to disappear subsequent to reduced N deposition [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of existant plant communities.

Recovery of existant plant communities from high N deposition would imply that the state of communities measured at different points in time is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness, a biodiversity endpoint that can be assessed relatively simple and communicated easily, is often negatively related to nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favourable changes can be derived from the traits of the species in a community. In Europe, environmental preference of plants has often been expressed using indicator values assigned to each plant species [@Ellenberg1992; @Landolt2010]. Examples of such metrics would be the number of oligotrophic species or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus a useful approach to understanding biodiversity change is through estimates of species turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction is compared to random expectation [@Chase2011].

In Switzerland, grassland accounts for 70% of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are also of agronomical importance [@Leiber2006]. In mountain hay meadows, the spatial variation of species richness in vascular plants is negative correlated with N deposition [@Roth2013], suggesting that mountain grasslands suffered from increased N deposition. However, between 1990 and 2010 NO$_x$ emissions in Switzerland decreased by 46% and NH$_3$ emissions by 14% [@MaasGrennfelt2016]. What may have led to a partial recovery of plant communities.

In addition to N deposition, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with N deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at the landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014]. Particularly the shift of plant communities at mountain summits are the result of recent climate warming and the aothors assume an interacting with airborne N deposition [@Steinbauer2018]. Furthermore, traditional management regimes are currently changing, which also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of accessible mountainous areas are frequently intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used data from the Swiss biodiversity monitoring [@Weber2004] to address the following questions: (1) Did biodiversity endpoints that are likely to reflect temperature, precipitation, N deposition or land-use intensity change over the last 15 years, (2) was species turnover correlated with the average temperature, precipitation, N deposition or inclination (we expect steep areas to be less intensively managed), and (3) did species that newly colonized or disappeared (local extinction) from sites differ from random expectation according to their indicator values for temperature, humidity, nutrients or light.

@@

## Monitoring data and community measures {-}
We analysed the presence/absence of vascular plants sampled within the scope of Switzerland's Biodiversity Monitoring (BDM) programme that was launched in 2001 to monitor Switzerland's biodiversity and to comply with the Convention on Biological Diversity of Rio de Janeiro [@Weber2004]. The sampling sites were circles with a size of 10 m$^2$ and data collection was carried out by qualified botanists who visited each sampling site twice within the same season. During each visit all the vascular plant species detected on the plot were recorded. For details on the field methods see @Plattner2004, @Roth2013 and @Roth2017.

After the sampling of the plant data the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008]. We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). We analysed the data from 2003 to 2017. During that study period each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012 and the third from 2013 to 2017. These selection criteria yielded a sample of `r nrow(sites)` sites.

For each survey (a survey consists of two  visits per season) we calculated the following biodiversity endpoints [@Rowe2017]: (1) The number of recorded species (species richness), the community mean of indicator values of recorded species for (2) temperature, (3) humidity, (4) nutrients and (5) light. The indicator values were obtained from the recalibrated indicator values for the Swiss Flora [@Landolt2010]. In addition to these five biodiversity endpoints that describe the state of plant communities for each site at a given time point, we also estimated the temporal turnover (i.e. species exchange ratio sensu @Hillebrand2018) as the proportion of species that differ between two time points to describe the community change between two points in time.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First, we expected communities to response to climate warming with this response being dependent on the position of the site along the temperature gradient [@Roth2014]. To describe the temperature gradient, we used the mean annual temperature per site from the WordClim database [@Fick2017].
The average $\pm$ SD mean annual temperature at our sites was `r mean(sites$TemperatureR / 10)` $\pm$ `r sd(sites$TemperatureR / 10)` $^\circ$C. 
 From the same database we used the annual precipitation per site, another key driver for plants that is likely to be affected by climate change  [@Beier2012].The average annual precipitation at our sites was `r mean(sites$PrecipitationR)` $\pm$ `r sd(sites$PrecipitationR)` mm. 
 Further, we estimated atmospheric N deposition for each site using a pragmatic approach  that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models [@RihmAchermann2016].
The average nitrogen deposition at our sites was `r mean(sites$NTOT_2000)` $\pm$ `r sd(sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ in 2000 and `r mean(sites$NTOT_2015)` $\pm$ `r sd(sites$NTOT_2015)` kg ha$^{-1}$ yr$^{-1}$ in 2015. 
Finally, we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015]. The average inclination at our sites was `r mean(sites$InclinationR)` $\pm$ `r sd(sites$InclinationR)`$^\circ$. 
## Statistical analyses {-}
To estimate the linear trend over time for each of the five biodiversity endpoints we applied linear mixed models (LMM) with normal distribution except for species-richness with Poisson distribution. We specified site specific trends with the assumption that the between-site differences in intercepts and slopes can be described with normal distributions [i.e. a random intercept random slope model, @Gelman2006]. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

To infer whether species turnover was changing along the gradient, we used a Binomial-GLMM with the proportion of species that differed between two surveys as dependent variable and the site gradients and period (first/second vs. second/third surveys) as predictors and site-ID as random effect. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

To infer whether species that colonized or disappeared from a site had particular indicator values that differed from the other species at that site, we produced for each site a list with all species that were recorded during the three surveys (total community). We then calculated the community mean (CM) of the indicator value for all species that colonized the site during the three surveys (i.e. not recorded during first survey and recorded during second or not recorded during second and recorded during third). We then randomly selected the same number of species from the total community and also calculated the community mean of the value for these species (random-CM). We repeated the random selection of species 1000 times. Then we calculated the differences of the CM minus the average of the random-CMs to obtain a standardized difference (standardized-CM) of how different the colonizing species were from random expectation. For example, a difference < 0 would suggest that the indicator values of colonizing species were lower than might be expected from random colonisation from the species-pool for this site. We applied this method for both colonizing and disappearing species and for the indicator values for temperature, humidity, nutrients and light (see Appendix A). We then tested whether this standardized difference is changing along the corresponding gradient (inferred from independet datasets, see Environmental gradients section) using linear models. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

We used logistic-GLMM to analyse whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this change depended on the species indicator value for nutrients. To analyse the colonization probability of species, we selected all species that were not observed during the first survey. For each of these species we asked if they were observed (then Occ = 1) or not observed (then Occ = 0) during the second survey. The same was also done for all species that were not observed during the second survey if they were observed or not observed (then Occ = 0) during the third survey. The variable 'occ' was then used as dependent variable in the logistic-GLMM. As predictor variables the model contained the N deposition of the site, the indicator value for nutrients of the species and the interaction of these two variables. Additionally, species-ID and site-ID were included as random effects. The same logistic-GLMM was also used to investigate local survival probabilities. In that case, however, we selected all species that were recorded during the first or second survey and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

To estimate the effect of N deposition on total species richness at a given point in time, we described the plant species richness at the sites using a generalized linear model with Poisson distribution and the logarithm as link function. As predictors we used the four environmental gradients as described above. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

@@

# Results {-}
## Temporal change in community structures {-}
```{r communitytrendscalc}
res <- tibble(Measures = c("Species-richness", "Temperature value", "Humidity value", "Nutrients value", "Light value"),
              'Period 1' = NA,
              'Period 2' = NA,
              'Period 3' = NA,
              'Trend' = NA,
              '5%' = NA,
              '95%' = NA,
              'Prob. for Trend' = NA)

@@

```{r communitytrendstab, echo=FALSE}
options(knitr.kable.NA = '') 
kable(res, align = c("l", rep("r", 7)), digits = 2,  booktabs = T, col.names = c("Measures", "Period 1", "Period 2", "Period 3", "Trend", "5%", "95%", "Prob. for trend"), caption = "Average measures of the biodiversity endpoints for the three sampling periods (period 1: 2003-2007; period 2: 2008-2012; period: 2013-2017). The temporal trends are given as change per 10 years and were estimated from linear mixed models with normal distribution (except for species richness with Poisson distribution and a log-link function). The measure of precision for the temporal trend is given as the 5\\% and 95\\% quantiles of the marginal posterior distribution of the linear trend. The column 'Prob. for trend' gives the probability that the linear trend is $> 0$. Indicator values according to Landolt et al. (2010).")
```

Species richness and four measures of plant community structure according to Landolt indicator values (i.e. biodiversity endpoints) suggested that plant communities in mountain hay meadows were rather stable between 2003 and 2017 and did not show a clear increase or decrease over time (Table \@ref(tab:communitytrendstab)): for each of the three 5-year survey periods the averages of species-richness and the average indicator values for temperature, humidity, nutrients and light did not vary much among the three sampling periods and the estimated trends were rather small. Except for average indicator value for temperature, the 90% credible-interval of the temporal trend contained zero. The results from the linear mixed models suggest that a linear temporal change was most likely for the community mean of the indicator value for temperature (probability of increase: `r 
res %>% filter(Measures == "Temperature value") %>% pull("Prob. for Trend")`), followed by the community mean of the indicator value for light (probability of decrease: `r 1 - res %>% filter(Measures == "Light value") %>% pull("Prob. for Trend")`). A linear temporal change was least likely for the species richness (probability of increase: `r res %>% filter(Measures == "Species-richness") %>% pull("Prob. for Trend")`). The chance that the community mean of the nutrient value decreased between 2003 and 2017 was `r 1- res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")` (`r res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")`gives the probability for an increase).

@@

```{r turnovertab, echo=FALSE}
options(knitr.kable.NA = '')
kable(res, align = c("l", rep("r", 3)), digits = 2,  booktabs = T, col.names = c("Gradient", "Estimate", "5%", "95%"), caption = "Change of species turnover along the four gradients. The slopes along the gradients (estimate) are given as the change per 10 years of the logit-probability of species that differed between two surveys. Estimates and the 5\\% and 95\\% quantiles of the marginal posterior distribution obtained from a Binomial-GLMM.")
```

The average $\pm$ SD percentage of species that differed between replicated surveys was `r mean(100 * turnover_double$turnover)` $\pm$ `r sd(100 * turnover_double$turnover)`%. This pseud-turnover was lower than the observed temporal turnover: the average percentage of species that differed between the first and second survey at a site was `r mean(100 * sites$change1 / (sites$nochange1 + sites$change1))` $\pm$ `r sd(100 * sites$change1 / (sites$nochange1 + sites$change1))`% and the percentage of species that differ between the second and third survey was `r mean(100 * sites$change2 / (sites$nochange2 + sites$change2))` $\pm$ `r sd(100 * sites$change2 / (sites$nochange2 + sites$change2))`%. Thus, it seemed that the turnover from the first/second survey to the turnover of the second/third survey moderately decreased (90% Credible interval of the change in turnover estimated from the Binomial generalized linear mixed model: `r CrI`). Along the N deposition gradient variation in species turnover was largest, with highest species turnover at sites with low N deposition (Table \@ref(tab:turnovertab)). The other three gradients were less conclusive to explain the variation in species turnover among sites.

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Indicator value", "Gradient", "Estimate", "5%", "90%", "Estimate", "5%", "90%"), caption = "Difference in the average indicator value of species that (a) disappeared from site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear model with the difference between disappeared/colonized species and random species as dependent variable and the site-measure (gradient) as predictor variable.") %>% 
  kable_styling() %>% 
  group_rows("(a) Plants that disappeard from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

>>>>

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
  - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity in many habitats. The recent introduction of cleaner technologies in Switzerland has led to a reduction in the emissions of nitrogen oxides, with an according decrease in N deposition. We examined different drivers of plant community change, i.e. N deposition, climate warming, and land-use change, in Swiss mountain hay meadows, using data from the Swiss biodiversity monitoring. We compared indicator values of species that disappeared from or colonized a site (species turnover) with the indicator values of randomly chosen species from the same site. While oligotrophic plant species were more likely to colonize, compared to random expectation, we found only weak shifts in plant community structure. In particular, the average nutrient value of plant communities remained stable over time (2003‚Äì2017). However, we found that the nutrient values of colonizing species showed the largest deviations from random expectation, suggesting that N deposition or other factors that change the nutrient content of soils were important drivers of the species composition change over the last 15 years in Swiss mountain hay meadows. In addition, we observed an overall replacement of species with lower indicator values for temperature with species with higher values. Apparently, the community effects of the replacement of eutrophic species with oligotrophic species was outweighed by climate warming. Our results add to the increasing evidence that inferring species turnover will generate a far more reliable understanding of the biotic response to changing environments than solely tracking average community composition.

@@

# Introduction {-}
Nitrogen (N) deposition is the entry of reactive nitrogen compounds into soil, water, and vegetation, input from the atmosphere to the biosphere. Since nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low nitrogen availability [@Vitousek1997], the addition of nitrogen is likely to change these communities. Indeed, together with land-use and climate change, N deposition is one of the major threats to biodiversity [@Sala2000; @Bobbink2010; @Murphy2013]. While there is strong evidence for the reduction in diversity of species-rich grasslands due to increased N deposition [@Stevens2006; @Dupre2010; @Maskell2010; @Wesche2012], mountain grasslands have received less attention [@Humbert2016].

In many parts of Europe, measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides since the late 1980s, with an according decrease in N deposition [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that suffered from increased N deposition. First, N deposition is still high at many sites, since in contrast to nitrogen oxides, there was on a minor recent decrease of ammonia emissions. Thus, even if N deposition is reduced, large areas might still be above the critical load [@Slootweeg2015; @RihmAchermann2016]. Second, there is a possibility that communities reach an alternative stable state after decades of increased N deposition and that the respective plant species are unlikely to disappear even if N deposition is reduced [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of plant communities.

Recovery of existing plant communities after high N deposition would imply that the state of communities measured at different points in time is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness, a biodiversity endpoint that can be assessed and communicated relatively easily, is often negatively related to nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favorable changes can be derived from the traits of the species in a community. In Europe, environmental preference of plants has often been expressed using indicator values assigned to each plant species [@Ellenberg1992; @Landolt2010]. Examples of such metrics would be the number of oligotrophic species, or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus, a useful approach to understanding biodiversity change is through estimates of species turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction is compared to random expectation [@Chase2011].

In Switzerland, grassland accounts for 70% of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are also of agronomical importance [@Leiber2006]. In mountain hay meadows, the spatial variation of species richness in vascular plants is negatively correlated with N deposition [@Roth2013], suggesting that mountain grasslands are negatively affected by increased N deposition. However, between 1990 and 2010, NO$_x$ emissions in Switzerland decreased by 46% and NH$_3$ emissions by 14% [@MaasGrennfelt2016]. Potentially, this could have led to a partial recovery of plant communities.

In addition to N deposition, Swiss mountain ecosystems are threatened also by other drivers of global change. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to to interact with N deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that, at the landscape scale, plant communities responded to climate warming even within a relatively short time period [@Roth2014]. [@Steinbauer2018] say that particularly the shift of plant communities at mountain summits are the result of recent climate warming, and they assume an interaction with airborne N deposition. Furthermore, traditional management regimes are currently changing, which also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of easily accessible mountainous areas are often being intensified, while poorly accessible mountainous areas are abandoned [@Tasser2002; @Strebel2015].

Here, we used data from the Swiss biodiversity monitoring [@Weber2004] to address the following questions: (1) Did biodiversity endpoints that are likely to reflect temperature, precipitation, N deposition, or land-use intensity change over the last 15 years? (2) Was species turnover correlated with the average temperature, precipitation, N deposition, or inclination (we expect steep areas to be less intensively managed)? (3) Did species that newly colonized or disappeared from (local extinction) sites differ from random expectation, according to their indicator values for temperature, humidity, nutrients, or light?

@@

## Monitoring data and community measures {-}
We analyzed the presence/absence of vascular plants sampled in the Swiss Biodiversity Monitoring (BDM) program that was launched in 2001 to monitor Switzerland's biodiversity and to comply with the Convention on Biological Diversity of Rio de Janeiro [@Weber2004]. The sampling sites were circles with a size of 10 m$^2$, and data collection was carried out by qualified botanists who visited each sampling site twice within the same season. During each visit, all vascular plant species detected on the plot were recorded. For details on the field methods see @Plattner2004, @Roth2013 and @Roth2017.

After the sampling of the plant data, the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008]. We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). We analyzed the data from 2003 to 2017. During that study period, each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012, and the third from 2013 to 2017. These selection criteria yielded a sample of `r nrow(sites)` sites.

For each survey (that consisted of two visits per season) we calculated the following biodiversity endpoints [@Rowe2017]: (1) The number of recorded species (species richness), the community mean of indicator values of recorded species for (2) temperature, (3) humidity, (4) nutrients, and (5) light. The indicator values were obtained from the recalibrated indicator values for the Swiss Flora [@Landolt2010]. In addition to these five biodiversity endpoints that describe the state of plant communities for each site at a given time point, we also estimated the temporal turnover (i.e. species exchange ratio sensu @Hillebrand2018) as the proportion of species that differ between two time points, to describe the community change between two points in time.

@@

## Environmental gradients {-}
We expected different drivers of global change to cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms, we ordered the sites along four main environmental gradients. First, we expected communities to response to climate warming [@Roth2014]. To describe the temperature gradient, we used the mean annual temperature per site from the WordClim database [@Fick2017].
The average $\pm$ SD mean annual temperature at our sites was `r mean(sites$TemperatureR / 10)` $\pm$ `r sd(sites$TemperatureR / 10)` $^\circ$C. 
 From the same database we used the annual precipitation per site, another key driver for plants that is likely to be affected by climate change [@Beier2012].The average annual precipitation at our sites was `r mean(sites$PrecipitationR)` $\pm$ `r sd(sites$PrecipitationR)` mm. 
 Further, we estimated atmospheric N deposition for each site using a pragmatic approach that combined monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models [@RihmAchermann2016].
The average nitrogen deposition at our sites was `r mean(sites$NTOT_2000)` $\pm$ `r sd(sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ in 2000 and `r mean(sites$NTOT_2015)` $\pm$ `r sd(sites$NTOT_2015)` kg ha$^{-1}$ yr$^{-1}$ in 2015. 
Finally, we used inclination as a proxy for land-use intensity, because we assumed that steeper sites are likely to be less intensively managed [@Strebel2015]. The average inclination at our sites was `r mean(sites$InclinationR)` $\pm$ `r sd(sites$InclinationR)`$^\circ$. 
## Statistical analyses {-}
To estimate the linear trend over time for each of the five biodiversity endpoints, we applied linear mixed models (LMM) with normal distribution except for species-richness with Poisson distribution. We specified site-specific trends with the assumption that the between-site differences in intercepts and slopes can be described with normal distributions [i.e. a random intercept random slope model, @Gelman2006]. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

To infer whether species turnover was changing along the gradient, we used a binomial generalized linear mixed model (GLMM) with the proportion of species that differed between two surveys as dependent variable and the site gradients and period (first/second vs. second/third surveys) as predictors, and with site-ID as random effect. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

To infer whether species that colonized or disappeared from a site had particular indicator values that differed from the other species at that site, we produced for each site a list with all species that were recorded during the three surveys (total community). We then calculated the community mean (CM) of the indicator value for all species that colonized the site during the three surveys (i.e. not recorded during the first survey and recorded during the second, or not recorded during the second and recorded during the third survey). We then randomly selected the same number of species from the total community and also calculated the community mean of the value for these species (random-CM). We repeated the random selection of species 1000 times. We then calculated the differences of the CM minus the average of the random-CMs to obtain a standardized difference (standardized-CM) of how different the colonizing species were from random expectation. For example, a difference < 0 would suggest that the indicator values of colonizing species were lower than might be expected from random colonization from the species-pool for this site. We applied this method for both colonizing and disappearing species and for the indicator values for temperature, humidity, nutrients, and light (see Appendix A). We then tested whether this standardized difference was changing along the corresponding gradient (inferred from independent datasets, see Environmental gradients section) using linear models. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

We used logistic GLMM to analyze whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this change depended on the species indicator value for nutrients. To analyze the colonization probability of species, we selected all species that were not observed during the first survey. For each of these species we asked if they were observed (Occ = 1) or not observed (Occ = 0) during the second survey. The same was also done for all species that were not observed during the second survey if they were observed or not observed during the third survey. The variable 'occ' was then used as dependent variable in the logistic GLMM. As predictor variables, the model contained the N deposition of the site, the indicator value for nutrients of the species, and the interaction of these two variables. Additionally, species-ID and site-ID were included as random effects. The same logistic GLMM was also used to investigate local survival probabilities. In that case, however, we selected all species that were recorded during the first or second survey, and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

To estimate the effect of N deposition on total species richness at a given point in time, we described the plant species richness at the sites using a generalized linear model (GLM) with Poisson distribution and the logarithm as link function. As predictors we used the four environmental gradients as described above. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

@@

# Results {-}
## Temporal change in community structures {-}
```{r communitytrendscalc}
res <- tibble(Measures = c("Species richness", "Temperature value", "Humidity value", "Nutrients value", "Light value"),
              'Period 1' = NA,
              'Period 2' = NA,
              'Period 3' = NA,
              'Trend' = NA,
              '5%' = NA,
              '95%' = NA,
              'Prob. for Trend' = NA)

@@

```{r communitytrendstab}
options(knitr.kable.NA = '') 
kable(res, align = c("l", rep("r", 7)), digits = 2,  booktabs = T, col.names = c("Measures", "Period 1", "Period 2", "Period 3", "Trend", "5%", "95%", "Prob. for trend"), caption = "Average measures of the biodiversity endpoints for the three sampling periods (period 1: 2003-2007; period 2: 2008-2012; period: 2013-2017). The temporal trends are given as change per 10 years and were estimated from linear mixed models with normal distribution (except for species richness with Poisson distribution and a log-link function). The measure of precision for the temporal trend is given as the 5\\% and 95\\% quantiles of the marginal posterior distribution of the linear trend (90\\% credible intervals). The column 'Prob. for trend' gives the probability that the linear trend is $> 0$. Indicator values according to Landolt et al. (2010).")
```

Species richness and the four measures of plant community structure according to Landolt indicator values (i.e. biodiversity endpoints) suggested that plant communities in mountain hay meadows were rather stable between 2003 and 2017 and did not show a clear increase or decrease over time (Table \@ref(tab:communitytrendstab)): for each of the three 5-year survey periods, the averages of species richness and the average indicator values for temperature, humidity, nutrients and light did not vary much among the three sampling periods, and the estimated trends were rather small. The results from the linear mixed models suggest that a linear temporal change was most likely for the community mean of the indicator value for temperature (probability of increase: `r 
res %>% filter(Measures == "Temperature value") %>% pull("Prob. for Trend")`), followed by the community mean of the indicator value for light (probability of decrease: `r 1 - res %>% filter(Measures == "Light value") %>% pull("Prob. for Trend")`). A linear temporal change was least likely for the species richness (probability of increase: `r res %>% filter(Measures == "Species-richness") %>% pull("Prob. for Trend")`). The chance that the community mean of the nutrient value decreased between 2003 and 2017 was `r 1- res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")` (`r res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")` gives the probability for an increase).

@@

```{r turnovertab, echo=FALSE}
options(knitr.kable.NA = '')
kable(res, align = c("l", rep("r", 3)), digits = 2,  booktabs = T, col.names = c("Gradient", "Estimate", "5%", "95%"), caption = "Change of species turnover along the four gradients. The slopes along the gradients (estimate) are given as the change per 10 years of the logit-probability of species that differed between two surveys. Estimates and 5\\% and 95\\% quantiles of the marginal posterior distribution obtained from a binomial GLMM.")
```

The average $\pm$ SD percentage of species that differed between replicated surveys was `r mean(100 * turnover_double$turnover)` $\pm$ `r sd(100 * turnover_double$turnover)`%. This turnover between replicated surveys was lower than the observed temporal turnover: the average percentage of species that differed between the first and second survey at a site was `r mean(100 * sites$change1 / (sites$nochange1 + sites$change1))` $\pm$ `r sd(100 * sites$change1 / (sites$nochange1 + sites$change1))`%, and the percentage of species that differed between the second and third survey was `r mean(100 * sites$change2 / (sites$nochange2 + sites$change2))` $\pm$ `r sd(100 * sites$change2 / (sites$nochange2 + sites$change2))`%. Thus, it seemed that the turnover from the first/second survey to the turnover of the second/third survey moderately decreased (90% credible interval of the change in turnover estimated from the binomial GLMM: `r CrI`). Along the N deposition gradient, variation in species turnover was largest, with highest species turnover at sites with low N deposition (Table \@ref(tab:turnovertab)). The other three gradients were less conclusive to explain the variation in species turnover among sites.

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Indicator value", "Gradient", "Estimate", "5%", "90%", "Estimate", "5%", "90%"), caption = "Difference in the average indicator value of species that (a) disappeared from a site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear models, with the difference between disappeared/colonized species and random species as dependent variable and the site-measure (gradient) as predictor variable.") %>% 
  kable_styling() %>% 
  group_rows("(a) Plants that disappeard from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

High species turnover at a site is the result of species that disappeared from the site and/or of species that newly colonized the site. To better understand the factors that drive these changes we were particularly interested whether the species that disappeared or colonized the sites differed in indicator values compared to what would be expected if the same number of species randomly disappeared or colonized the sites (i.e. random disappearance and random colonization), and whether there is spatial variation along the environmental gradients. It seems that the indicator values of newly colonizing species differed more from random colonization than the indicator values of disappearing species (Table \@ref(tab:difffromrandomtab)). For colonizing species, we found the largest differences from random colonization in the indicator value for nutrients: at sites with relatively low N deposition of 10 kg ha$^{-1}$ yr$^{-1}$, the newly colonizing species had on average a lower indicator value for nutrients than species under random colonization (column "Difference from random" in Table \@ref(tab:difffromrandomtab)), but the differences between colonizing species and random colonization decreased with increasing N deposition (column "Change along gradient" in Table \@ref(tab:difffromrandomtab)). Thus, at high N deposition, colonizing species did not differ from random species (see Figure 3b in Appendix A).
